Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-11 main.tex
Make some small changes

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer sience. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in particular\cite{paulson:isabelle}\todo{find references}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

In this paper we use the sequent calculus $\FOLK$, shown in
Figure~\ref{fig:LK1}, which is a one-sided variant of Gentzen's
original calculus~\cite{gentzen:35:I} for first-order logic. To
simplify some technicalities later in this paper, we also include here
the $\mixr$-rule.

@@

A \bfit{directed graph} $\gG =\tuple{\vG,\eG}$ is a set $\vG$ of
\bfit{vertices} and a set $\eG\subseteq \vG\times\vG$ of \bfit{direct edges}.
A \bfit{directed graph homomorphism}
$\phi\colon\tuple{\vG,\eG}\to\tuple{\vH,\eH}$ is function $\phi\colon\vG\to\vH$ such that if
$\pair{v,w}\in\eG$ then $\pair{\phi(v),\phi(w)}\in\eH$.

@@

In a graph $\gG$, the \bfit{neighbourhood} $N(v)$ of a vertex $v \in\vG$ is defined as
the set $\set{w \mid vw \in\eG}$. A \bfit{module} is a set $M \subseteq \vG$ such that $N(v) \ M = N(w) \ M$
for all $v, w \in M$. A module $M$ is \bfit{strong} if for every
module $M'$, we have $M' \subseteq M$, $M \subseteq$ or $M \cap M' =
\emptyset$. A module is \bfit{proper} if it has two or more vertices.

@@

\begin{proof}
  That  $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}. 
  \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}

@@

Figure~\ref{fig:KS1} shows the inference rule for the deep inference
system~$\FOKS$ that we are using this paper. It is a slight variation
of the systems presented by Br\"unnler~\cite{brunnler:phd} and
Ralph~\cite{ralph:phd} in their PhD-theses. The main differences being
that we (i) the explicit presence of the $\mixr$-rule, (ii) a
different choice of how the formula equivalence $\fequ$ is defined,
and (iii) an explicit rule for the equivalence.

>>>>

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer sience. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in
particular\cite{DSAV:15lean,paulson:isabelle}\todo{find
references}\juihsuan{done}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

In this paper we use the sequent calculus $\FOLK$, shown in
Figure~\ref{fig:LK1}, which is an one-sided variant of Gentzen's
original calculus~\cite{gentzen:35:I} for first-order logic. To
simplify some technicalities later in this paper, we also include here
the $\mixr$-rule.

@@

A \bfit{directed graph} $\gG =\tuple{\vG,\eG}$ is a set $\vG$ of
\bfit{vertices} and a set $\eG\subseteq \vG\times\vG$ of \bfit{direct edges}.
A \bfit{directed graph homomorphism}
$\phi\colon\tuple{\vG,\eG}\to\tuple{\vH,\eH}$ is a function $\phi\colon\vG\to\vH$ such that if
$\pair{v,w}\in\eG$ then $\pair{\phi(v),\phi(w)}\in\eH$.

@@

In a graph $\gG$, the \bfit{neighbourhood} $N(v)$ of a vertex $v \in\vG$ is defined as
the set $\set{w \mid vw \in\eG}$. A \bfit{module} is a set $M \subseteq \vG$
such that $N(v) \setminus M = N(w) \setminus M$
for all $v, w \in M$. A module $M$ is \bfit{strong} if for every
module $M'$, we have $M' \subseteq M$, $M \subseteq$ or $M \cap M' =
\emptyset$. A module is \bfit{proper} if it has two or more vertices.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
structural induction on $A$:
  \begin{itemize}
    \item Trivial if $A$ is an atom.
    \item The induction step uses the following fact:
	  if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
          module of $\gG$ if and only if it is a module of $\gG[X]$.
  \end{itemize}
  \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
\juihsuan{I'm still looking for a simple proof}
\footnote{Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then $\graphof A = \single x \quad
\single p(x) \quad \single x \quad \single q(x)$, the scope of each $x$-binder
contains all the vertices, in particular, the two $x$-binders.}

@@

Figure~\ref{fig:KS1} shows the inference rules for the deep inference
system~$\FOKS$ that we are using in this paper. It is a slight variation
of the systems presented by Br\"unnler~\cite{brunnler:phd} and
Ralph~\cite{ralph:phd} in their PhD-theses. The main differences being
that we have (i) the explicit presence of the $\mixr$-rule, (ii) a
different choice of how the formula equivalence $\fequ$ is defined,
and (iii) an explicit rule for the equivalence.
